---
title: "Workflow: Methylome Preprocessing, Epimutation-Rate Estimation and QTL Mapping"
author: "Zhilin Zhang"
date: "2026-05"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: cosmo
    code_folding: show
params:
  run_code: false
  project_dir: "/path/to/project"
  github_repo: "https://github.com/zhilinzhang340/Mapping-genetic-modifiers-of-epimutation-rates-codes-Zhilin_Zhang"
  zenodo_doi: "To be updated upon Zenodo release"
  alphabeta_job_table: "metadata/alphabeta_jobs.tsv"
  alphabeta_out_dir: "results/02_alphabeta"
  qtl_data_dir: "data/qtl_cross_files"
  qtl_out_dir: "results/03_qtl_mapping"
  newbuildPedigree_script: "newbuildPedigree_stand_alone_rmCM_2023_11.R"
  mCG_cross_file: "newrun_2023-11_68_RIL_all_annotations_mCG_level_without_ID_SNP_markers_after_using_SNP_sliding_windows_cleaning_2025-04.csv"
  alpha_cross_file: "newrun_2023-11_68_RIL_all_annotations_add_intergeneic_alpha_rate_without_ID_SNP_markers_after_using_SNP_sliding_windows_cleaning_2025-04.csv"
  beta_cross_file: "newrun_2023-11_68_RIL_all_annotations_add_intergeneic_beta_rate_without_ID_SNP_markers_after_using_SNP_sliding_windows_cleaning_2025-04.csv"
  qtl_traits: !r c("global", "gene", "TE", "gbM", "UM", "teM", "intergenic")
  qtl_error_prob: 1e-7
  qtl_n_perm: 1000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11,
  fig.height = 5
)
set.seed(1)
```

# Overview

This R Markdown notebook documents the complete computational workflow for:

> **"Mapping genetic modifiers of epimutation rates identifies VIM2/4 as dosage-sensitive negative regulators of CG methylation maintenance"** (Zhang et al.)

The pipeline consists of three main steps:

1. **MethylStar**: WGBS read alignment, methylation extraction, and Methimpute methylation-state calling
2. **AlphaBeta**: Pedigree-based estimation of CG methylation gain rate (alpha) and loss rate (beta)
3. **R/qtl**: QTL mapping of methylation levels and epimutation rate variation across 68 Ler x Cvi RILs

By default, `params$run_code` is set to `FALSE` so the document can be knitted without access to the full dataset. To rerun the analysis, set `run_code: true` in the YAML header and update all paths under `params`.

## Code and data resources

| Resource | Description | Link |
|:---------|:------------|:-----|
| MethylStar | WGBS preprocessing, Bismark alignment, Methimpute calling | <https://github.com/jlab-code/MethylStar> |
| AlphaBeta | Pedigree-based epimutation rate estimation | <https://github.com/jlab-code/AlphaBeta> |
| Project code | Analysis scripts and this workflow notebook | `r params$github_repo` |
| WGBS data | 371 MAML samples (GEO) | [GSE296957](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE296957) |

## Expected input structure

| File or directory | Description |
|:------------------|:------------|
| `data/raw_fastq/` | Raw paired-end WGBS FASTQ files |
| `metadata/sample_metadata_index.tsv` | Sample metadata (sample IDs, RIL IDs, lineage IDs, genotype background, generation, file names) |
| `metadata/alphabeta_jobs.tsv` | AlphaBeta job table mapping each run to its node list, edge list, annotation class and output prefix |
| `data/qtl_cross_files/*.csv` | R/qtl cross files with genetic markers and molecular traits |
| `newbuildPedigree_stand_alone_rmCM_2023_11.R` | Helper script for building AlphaBeta pedigree input |

# Step 1: WGBS Preprocessing and Methylation-State Calling (MethylStar)

[MethylStar](https://github.com/jlab-code/MethylStar) automates the full WGBS preprocessing pipeline: trimming, quality control, Bismark alignment against the TAIR10 reference genome, duplicate removal, methylation extraction, cytosine-call generation, and Methimpute methylation-state calling.

The final output is a set of per-sample Methimpute files containing cytosine-level methylation states classified as unmethylated (U), intermediate (I), or methylated (M), together with posterior probabilities. Only high-confidence calls (`posteriorMax >= 0.99`) were retained for downstream AlphaBeta rate estimation.

## 1.1 Installation

```{bash methylstar-install, eval=FALSE}
# Option A: install from GitHub
git clone https://github.com/jlab-code/MethylStar.git
cd MethylStar
python2 run.py
```

## 1.2 MethylStar run settings

Within the MethylStar menu:

- `C. Configuration`: set raw FASTQ directory, output directory, reference genome (TAIR10), and parallelization parameters.
- `1. Run Pipeline (WGBS)` > `A. Quick Run` > Option `0`: Trimmomatic, QC, Bismark alignment, duplicate removal, methylation extraction, cytosine calls, and Methimpute.


A total of 371 MAML samples across 68 Ler x Cvi RILs and two parental MA-lines (Ler-0 and Cvi-0) were processed.

# Step 2: Epimutation-Rate Estimation (AlphaBeta)

[AlphaBeta](https://github.com/jlab-code/AlphaBeta) estimates per-generation CG methylation gain rate (alpha) and loss rate (beta) from pedigree-structured, multi-generational methylation-state data.

## 2.1 Installation

```{r install-alphabeta, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("AlphaBeta")
```

## 2.2 Input files

Each RIL requires two input files:

**Nodelist** -- lists all samples in the pedigree with their methylome file paths:

| mfile | node | gen |
|:------|:-----|:----|
| /path/to/methylome_G02_L1.txt | G02_L1 | 2 |
| /path/to/methylome_G09_L1.txt | G09_L1 | 9 |

**Edgelist** -- defines parent-offspring relationships:

| from | to |
|:-----|:---|
| G02_L1 | G09_L1 |

## 2.3 AlphaBeta wrapper function

The wrapper performs four operations: (1) builds the pedigree object, (2) fits the neutral model (ABneutral), (3) fits the null model (ABnull) and compares by F-test, (4) bootstraps standard errors.

The original analysis was implemented using the following scripts:

- `run_AB_global_rate_check_samples_RIL4.R`
- `function_run_alphabeta_nstarts_2000_CG_progenitor_intermediate_without_selection_using_newp0uu_rmCM_2023-11.R`
- `newbuildPedigree_stand_alone_rmCM_2023_11.R`



```{r alphabeta-wrapper, eval=params$run_code}
library(AlphaBeta)
library(data.table)

run_alphabeta_job <- function(nodelist,
                              edgelist,
                              job_name,
                              input_dir,
                              output_dir,
                              newbuild_pedigree_script,
                              cytosine = "CG",
                              posterior_max_filter = 0.99,
                              n_starts = 2000,
                              n_boot = 1000) {
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
  source(newbuild_pedigree_script)

  path_node <- file.path(input_dir, nodelist)
  path_edge <- file.path(input_dir, edgelist)
  if (!file.exists(path_node)) stop("Node list not found: ", path_node)
  if (!file.exists(path_edge)) stop("Edge list not found: ", path_edge)

  old_wd <- getwd()
  on.exit(setwd(old_wd), add = TRUE)
  setwd(output_dir)

  # --- Visualize pedigree ---
  plotPedigree(
    nodelist = path_node, edgelist = path_edge,
    sampling.design = "progenitor.intermediate",
    output.dir = output_dir,
    plot.width = 8, plot.height = 8,
    aspect.ratio = 0.5, vertex.size = 4, vertex.label = TRUE,
    out.pdf = paste0("pedigree_", job_name)
  )

  # --- Build pedigree from methylome files ---
  ab_input <- newbuildPedigree(
    path_node, path_edge,
    cytosine = cytosine,
    posteriorMaxFilter = posterior_max_filter
  )

  p0uu_in  <- ab_input$tmpp0
  pedigree <- ab_input$Pdata

  fwrite(data.table(p0uu = p0uu_in),
         file.path(output_dir, paste0("p0uu_in_", job_name, ".tsv")), sep = "\t")
  fwrite(as.data.table(pedigree),
         file.path(output_dir, paste0("pedigree_pdata_", job_name, ".tsv")), sep = "\t")

  # --- ABneutral model ---
  output_ABneutral <- ABneutral(
    pedigree.data = pedigree,
    p0uu = p0uu_in, eqp = p0uu_in, eqp.weight = 1,
    Nstarts = n_starts,
    out.dir = output_dir,
    out.name = paste0("ABneutral_CG_estimates_", job_name)
  )

  ABplot(
    pedigree.names = paste0("ABneutral_CG_estimates_", job_name, ".Rdata"),
    output.dir = output_dir,
    out.name = paste0("ABneutral_CG_", job_name),
    plot.height = 8, plot.width = 11
  )

  neutral_estimates <- output_ABneutral$estimates[1:10, 1:5] |>
    as.data.frame() |>
    tibble::rownames_to_column("parameter") |>
    as.data.table()
  fwrite(neutral_estimates,
         file.path(output_dir, paste0("ABneutral_estimates_", job_name, ".tsv")), sep = "\t")

  # --- Diagnostic plot ---
  pdf(file.path(output_dir, paste0("ABneutral_pedigree_", job_name, ".pdf")),
      width = 11, height = 8)
  plot(output_ABneutral$pedigree[, "delta.t"],
       output_ABneutral$pedigree[, "div.obs"],
       xlab = "Divergence time (delta.t)",
       ylab = "Observed methylation divergence", main = job_name)
  points(output_ABneutral$for.fit.plot[, "delta.t"],
         output_ABneutral$for.fit.plot[, "div.sim"], type = "l")
  dev.off()

  # --- ABnull model and F-test ---
  output_ABnull <- ABnull(
    pedigree.data = pedigree,
    out.dir = output_dir,
    out.name = paste0("ABnull_CG_estimates_", job_name)
  )

  ftest <- FtestRSS(
    pedigree.select = paste0("ABneutral_CG_estimates_", job_name, ".Rdata"),
    pedigree.null   = paste0("ABnull_CG_estimates_", job_name, ".Rdata")
  )

  ftest_table <- data.table(job_name = job_name,
                            ftest_p_value = as.numeric(ftest$Ftest[6]))
  fwrite(ftest_table,
         file.path(output_dir, paste0("ABneutral_vs_ABnull_Ftest_", job_name, ".tsv")), sep = "\t")

  # --- Bootstrap standard errors ---
  boot_output <- BOOTmodel(
    pedigree.data = paste0("ABneutral_CG_estimates_", job_name, ".Rdata"),
    Nboot = n_boot,
    out.dir = output_dir,
    out.name = paste0("ABneutral_Boot_CG_estimates_", job_name)
  )

  boot_summary <- data.table(
    job_name           = job_name,
    alpha              = boot_output$boot.base$alpha,
    beta               = boot_output$boot.base$beta,
    alpha_se           = boot_output$standard.errors[1, 1],
    beta_se            = boot_output$standard.errors[2, 1],
    beta_alpha_ratio_se = boot_output$standard.errors[3, 1]
  )
  fwrite(boot_summary,
         file.path(output_dir, paste0("ABneutral_bootstrap_summary_", job_name, ".tsv")), sep = "\t")

  list(
    job_name = job_name, p0uu = p0uu_in, pedigree = pedigree,
    neutral = output_ABneutral, null = output_ABnull,
    ftest = ftest, bootstrap = boot_output, summary = boot_summary
  )
}
```

## 2.4 Batch run: all annotation-specific jobs

The same wrapper is applied to global CG sites and to annotation-specific subsets (gene, gbM, UM, TE, teM, intergenic) via a job table.

```{r expected-job-columns}
expected_alphabeta_job_cols <- c(
  "job_name", "cytosine_context", "annotation",
  "nodelist", "edgelist", "input_dir", "output_dir"
)
expected_alphabeta_job_cols
```

```{r run-alphabeta-jobs, eval=params$run_code}
job_table <- fread(file.path(params$project_dir, params$alphabeta_job_table))

missing_job_cols <- setdiff(expected_alphabeta_job_cols, names(job_table))
if (length(missing_job_cols) > 0) {
  stop("Missing AlphaBeta job-table columns: ", paste(missing_job_cols, collapse = ", "))
}

newbuild_script <- file.path(params$project_dir, params$newbuildPedigree_script)

alphabeta_results <- lapply(seq_len(nrow(job_table)), function(i) {
  x <- job_table[i]
  run_alphabeta_job(
    nodelist   = x$nodelist,
    edgelist   = x$edgelist,
    job_name   = x$job_name,
    input_dir  = file.path(params$project_dir, x$input_dir),
    output_dir = file.path(params$project_dir, x$output_dir),
    newbuild_pedigree_script = newbuild_script,
    cytosine = x$cytosine_context
  )
})
```

## 2.5 Standalone example: global CG rate estimation

```{r run-global-example, eval=FALSE}
run_alphabeta_job(
  nodelist   = "nodelist_update_RIL4.txt",
  edgelist   = "RILRaw_edgelist.txt",
  job_name   = "RIL4_global_rate",
  input_dir  = "data/pedigree/global/",
  output_dir = "results/02_alphabeta/global/",
  newbuild_pedigree_script = "newbuildPedigree_stand_alone_rmCM_2023_11.R",
  cytosine = "CG"
)
```

# Step 3: QTL Mapping

QTL scans were performed using the R/qtl package with Haley-Knott regression. The same workflow was applied three times for:

1. **Methylation-level QTL** (steady-state mCG levels, Fig. 2c)
2. **Alpha-rate QTL** (methylation gain rates, Fig. 3c)
3. **Beta-rate QTL** (methylation loss rates, Fig. 3c)

Genotypes are coded as `0` (Ler) and `2` (Cvi). The genetic map was re-estimated with `error.prob = 1e-7`. Genome-wide significance thresholds were determined from 1000 permutations. LOD confidence intervals were obtained with `lodint(drop = 2)`.

The original analysis was implemented using the following script:

- `plot_Figure3C.R`

## 3.1 QTL workflow function

```{r qtl-function, eval=params$run_code}
library(qtl)
library(data.table)
library(ggplot2)
library(scales)

run_single_qtl_workflow <- function(cross_file,
                                    traits,
                                    output_prefix,
                                    output_dir,
                                    error_prob = 1e-7,
                                    n_perm = 1000,
                                    lod_drop = 2,
                                    method = "hk") {
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
  if (!file.exists(cross_file)) stop("Cross file not found: ", cross_file)

  cross <- read.cross(file = cross_file, format = "csv", genotypes = c("0", "2"))
  cross_map <- est.map(cross, error.prob = error_prob)
  cross <- replace.map(cross, cross_map)
  cross <- calc.genoprob(cross, step = 1)

  scan_list   <- list()
  perm_list   <- list()
  peak_list   <- list()
  lodint_list <- list()

  for (trait in traits) {
    if (!trait %in% names(cross$pheno)) {
      warning("Trait not found: ", trait)
      next
    }

    # Single-QTL scan
    scan_res <- scanone(cross, pheno.col = trait, method = method)
    scan_dt  <- as.data.table(scan_res)
    scan_dt[, trait := trait]
    setnames(scan_dt, old = "lod", new = "LOD")
    scan_list[[trait]] <- scan_dt

    # Permutation threshold
    perm_res   <- scanone(cross, pheno.col = trait, method = method, n.perm = n_perm)
    cutoff_05  <- as.numeric(summary(perm_res)[1, 1])
    perm_list[[trait]] <- data.table(trait = trait, alpha = 0.05,
                                     LOD_threshold = cutoff_05, n_perm = n_perm)

    # Peak and LOD interval
    peak <- scan_dt[which.max(LOD)]
    peak[, LOD_threshold := cutoff_05]
    peak[, significant := LOD >= LOD_threshold]
    peak_list[[trait]] <- peak

    lodint_res <- tryCatch(
      as.data.table(lodint(scan_res, chr = peak$chr, drop = lod_drop)),
      error = function(e) data.table()
    )
    if (nrow(lodint_res) > 0) {
      lodint_res[, trait := trait]
      lodint_res[, lod_drop := lod_drop]
    }
    lodint_list[[trait]] <- lodint_res
  }

  scan_all   <- rbindlist(scan_list, fill = TRUE)
  perm_all   <- rbindlist(perm_list, fill = TRUE)
  peak_all   <- rbindlist(peak_list, fill = TRUE)
  lodint_all <- rbindlist(lodint_list, fill = TRUE)

  fwrite(scan_all,   file.path(output_dir, paste0(output_prefix, "_scanone_all.tsv")), sep = "\t")
  fwrite(perm_all,   file.path(output_dir, paste0(output_prefix, "_permutation_thresholds.tsv")), sep = "\t")
  fwrite(peak_all,   file.path(output_dir, paste0(output_prefix, "_QTL_peaks.tsv")), sep = "\t")
  fwrite(lodint_all, file.path(output_dir, paste0(output_prefix, "_LOD2_intervals.tsv")), sep = "\t")

  # LOD curve plot
  p <- ggplot(scan_all, aes(x = pos, y = LOD, color = trait)) +
    geom_line(linewidth = 0.45) +
    geom_hline(data = perm_all,
               aes(yintercept = LOD_threshold, color = trait),
               linetype = "dashed", linewidth = 0.25, show.legend = FALSE) +
    facet_wrap(~ chr, nrow = 1, scales = "free_x") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    labs(title = output_prefix, x = "Genetic position (cM)", y = "LOD", color = "Trait") +
    theme_classic(base_size = 14) +
    theme(strip.background = element_blank(), plot.title = element_text(hjust = 0.5))

  ggsave(filename = file.path(output_dir, paste0(output_prefix, "_LOD_curve.pdf")),
         plot = p, width = 11, height = 5)

  list(cross = cross, scan = scan_all, permutation = perm_all,
       peaks = peak_all, lodint = lodint_all, plot = p)
}
```

## 3.2 Methylation-level QTL (Fig. 2c)

Steady-state mCG levels across genomic compartments were mapped as molecular quantitative traits.

```{r qtl-mcg-level, eval=params$run_code}
qtl_data_dir <- file.path(params$project_dir, params$qtl_data_dir)
qtl_out_dir  <- file.path(params$project_dir, params$qtl_out_dir)

qtl_mCG <- run_single_qtl_workflow(
  cross_file    = file.path(qtl_data_dir, params$mCG_cross_file),
  traits        = params$qtl_traits,
  output_prefix = "Fig2c_mCG_level_QTL",
  output_dir    = file.path(qtl_out_dir, "Fig2c_mCG_level_QTL"),
  error_prob    = params$qtl_error_prob,
  n_perm        = params$qtl_n_perm
)
qtl_mCG$plot
```

## 3.3 Alpha-rate QTL (Fig. 3c)

AlphaBeta-estimated methylation gain rates (alpha) were mapped as quantitative traits.

```{r qtl-alpha-rate, eval=params$run_code}
qtl_alpha <- run_single_qtl_workflow(
  cross_file    = file.path(qtl_data_dir, params$alpha_cross_file),
  traits        = params$qtl_traits,
  output_prefix = "Fig3c_alpha_rate_QTL",
  output_dir    = file.path(qtl_out_dir, "Fig3c_alpha_rate_QTL"),
  error_prob    = params$qtl_error_prob,
  n_perm        = params$qtl_n_perm
)
qtl_alpha$plot
```

## 3.4 Beta-rate QTL (Fig. 3c)

AlphaBeta-estimated methylation loss rates (beta) were mapped as quantitative traits.

```{r qtl-beta-rate, eval=params$run_code}
qtl_beta <- run_single_qtl_workflow(
  cross_file    = file.path(qtl_data_dir, params$beta_cross_file),
  traits        = params$qtl_traits,
  output_prefix = "Fig3c_beta_rate_QTL",
  output_dir    = file.path(qtl_out_dir, "Fig3c_beta_rate_QTL"),
  error_prob    = params$qtl_error_prob,
  n_perm        = params$qtl_n_perm
)
qtl_beta$plot
```

# Output Summary

| Step | Analysis | Main outputs | Manuscript figure |
|:-----|:---------|:-------------|:------------------|
| 1 | WGBS preprocessing | Methimpute cytosine-state files; QC reports | Input for Steps 2-3 |
| 2 | AlphaBeta rate estimation | ABneutral estimates; F-test tables; bootstrap SEs | Fig. 3a,b |
| 3 | mCG level QTL | Scanone results; QTL peaks; LOD intervals; LOD-curve PDF | Fig. 2c |
| 3 | Alpha-rate QTL | Scanone results; QTL peaks; LOD intervals; LOD-curve PDF | Fig. 3c |
| 3 | Beta-rate QTL | Scanone results; QTL peaks; LOD intervals; LOD-curve PDF | Fig. 3c |

# Reproducibility Notes

- MethylStar was used for WGBS preprocessing and Methimpute methylation-state calling.
- AlphaBeta was used to estimate genotype-specific CG methylation gain and loss rates from pedigree-structured methylation-state data.
- QTL mapping was performed with the R package `qtl` (Haley-Knott regression).
- The genetic map was estimated with `est.map()` using `error.prob = 1e-7`.
- Genome-wide significance thresholds were estimated from 1000 permutations.
- LOD confidence intervals were obtained using `lodint(drop = 2)`.

# Session Information

```{r session-info}
sessionInfo()
```
